quollr: An R Package for Visualizing 2-D Models from Nonlinear Dimension Reductions in High-Dimensional Space

### 2025-12-19 paper-quollr.Rmd
Updated spellings and try to make submit .zip file

# # GitHub raw URL of the figure
# fig_url <- "https://raw.githubusercontent.com/JayaniLakshika/paper-quollr/main/figures/quollr_workflow.png"
# 
# # Determine figure path depending on output
# if (knitr::is_html_output()) {
#   # HTML can use the URL directly
#   fig_path <- fig_url
# } else {
#   # PDF needs a local copy
#   temp_file <- tempfile(fileext = ".png")
#   download.file(fig_url, destfile = temp_file, mode = "wb")
#   fig_path <- temp_file
# }
# 
# knitr::include_graphics(fig_path)

@@

```{r geom-outputs, fig.cap="The outputs of `geom_hexgrid` and `geom_trimesh` include: (a) a complete hexagonal grid, (b) a hexagonal grid that corresponds with the data, (c) a full grid based on centroid triangulation, and (d) a centroid triangulation grid that aligns with the data.", fig.align='center', fig.pos="!ht", fig.height=5, fig.width=20, eval=knitr::is_html_output(), fig.alt = "The figure consists of four 2-D panels illustrating hexagonal grids and triangulated meshes. Panel (a) shows a complete hexagonal grid covering a rectangular 2-D region, with all hexagons drawn regardless of whether data are present. Panel (b) shows a hexagonal grid restricted to bins that contain data points, with only those hexagons displayed and gaps where bins are empty. Panel (c) shows a full triangulated mesh constructed from centroids across the entire grid region, with centroids connected by straight line segments to form triangles. Panel (d) shows a triangulated mesh constructed only from centroids corresponding to non-empty bins, aligned with the data layout and leaving open areas where no centroids are present. In all panels, the horizontal and vertical axes represent continuous 2-D coordinates with bounded numeric ranges, and the grids and meshes are drawn with uniform line styles and colours."}

@@

```{r geom-outputs-pdf, fig.cap="The outputs of \\texttt{geom\\_hexgrid} and \\texttt{geom\\_trimesh} include: (a) a complete hexagonal grid, (b) a hexagonal grid that corresponds with the data, (c) a full grid based on centroid triangulation, and (d) a centroid triangulation grid that aligns with the data.", fig.align='center', fig.pos="!ht", fig.height=5, fig.width=20, eval=knitr::is_latex_output(), fig.alt = "The figure consists of four 2-D panels illustrating hexagonal grids and triangulated meshes. Panel (a) shows a complete hexagonal grid covering a rectangular 2-D region, with all hexagons drawn regardless of whether data are present. Panel (b) shows a hexagonal grid restricted to bins that contain data points, with only those hexagons displayed and gaps where bins are empty. Panel (c) shows a full triangulated mesh constructed from centroids across the entire grid region, with centroids connected by straight line segments to form triangles. Panel (d) shows a triangulated mesh constructed only from centroids corresponding to non-empty bins, aligned with the data layout and leaving open areas where no centroids are present. In all panels, the horizontal and vertical axes represent continuous 2-D coordinates with bounded numeric ranges, and the grids and meshes are drawn with uniform line styles and colours."}

@@

```{r scurve-highd-model-pdf, eval=knitr::is_latex_output(), fig.pos="!ht", fig.cap="$2\\text{-}D$ projections of the lifted high-dimensional wireframe model from the \\texttt{scurve} UMAP layout. Each panel (a1–a4) shows the model (black) overlaid on \\texttt{scurve} data (blue) in different projections. These views illustrate how the lifted wireframe model captures the structure of the \\texttt{scurve} data. Regions with sparse or no data in the UMAP layout are also visible in the lifted model.", fig.alt = "The figure shows four panels of 2-D scatter plots (a1–a4), each displaying a different projection of the lifted high-dimensional wireframe model derived from the scurve UMAP layout. In each panel, the horizontal and vertical axes represent two projected coordinates with continuous numeric ranges. The wireframe model is drawn in black as a wireframe of connected line segments, and the scurve data points are overlaid in blue as individual markers. The relative positions and orientations of the wireframe and data points vary across panels as different projections are shown. In some regions, the wireframe extends into areas with few or no data points, leaving visible gaps between the model and the data. All panels have similar aspect ratios and use consistent colour and symbol encodings to distinguish the model from the data."}

@@

```{r scurve-highd-model-detourrpdf, out.width=c("50%", "50%"), fig.cap="Screenshots of the lifted high-dimensional wireframe model from the \\texttt{scurve} UMAP layout using \\texttt{detourr}. Regions with sparse or no data in the UMAP layout are also visible in the lifted model.", fig.alt = "The figure shows several screenshots of 2-D projections of the lifted high-dimensional wireframe model derived from the scurve UMAP layout using detourr. In each screenshot, the horizontal and vertical axes represent projected coordinates with continuous numeric ranges. The wireframe model is displayed as a set of connected line segments, and the scurve data points are shown as individual markers overlaid on the wireframe. The orientation and shape of the wireframe and point cloud differ across screenshots as different projections are shown. In some views, parts of the wireframe extend into areas with few or no data points, leaving visible gaps between the model and the data. All screenshots use consistent colour and line styles and have similar aspect ratios.", fig.pos="H", fig.show='hold', eval=knitr::is_latex_output()}

@@

```{r scurve-nldrlink-html, echo=TRUE, eval=knitr::is_html_output(), fig.cap="Interactively linked plots connecting the UMAP layout (left) and the fitted model overlaid with the data in $7\\text{-}D$ (right) using `langevitour`. Brushing in either view highlights the corresponding subset of points across both representations, allowing examination of how well the UMAP layout preserves the underlying `scurve` structure. Double-clicking on either view clears the selection and dehighlights all points.", fig.alt = "The figure shows two interactively linked panels. The left panel displays a 2-D UMAP layout of the scurve data, with horizontal and vertical axes representing the UMAP coordinates on continuous numeric scales. The right panel shows a 2-D projection of the 7-D data and fitted model using langevitour, where each axis represents a linear combination of the original seven dimensions. Data points are plotted as individual markers, and brushing a subset of points in either panel highlights the corresponding points in the other panel using a distinct colour. The highlighted points update dynamically as selections change, and double-clicking on either panel clears all highlights. The display maintains roughly square aspect ratios in both panels and uses consistent visual encodings for data and selected subsets throughout the interactive view.", layout = "l-body"}

@@

```{r scurve-nldrlink-pdf, eval=knitr::is_latex_output(), fig.pos="!ht", fig.width=10, fig.height=10, fig.cap="Exploring the correspondence between UMAP layout and \\texttt{scurve} structure in $7\\text{-}D$. Two sets of plots are interactively linked: UMAP layout (a1, b1) and projection of $7\\text{-}D$ model and data (a2, b2). The purple points indicate the selected subsets, which differ between rows. In (a1), the lower bridge of the \\texttt{scurve} is highlighted, which corresponds in (a2) to points spanning across both arms of the high-dimensional structure. In (b1), a different region near the upper arm of the \\texttt{scurve} is selected, and in (b2) these points map onto one side of the curved manifold in $7\\text{-}D$ projection. While the UMAP layout suggests distinct local clusters, the linked tour views reveal how these selections trace continuous structures in the $7\\text{-}D$ space, highlighting distortions introduced by UMAP.", fig.alt = "The figure consists of four linked panels arranged in two rows and two columns. Panels (a1) and (b1) show 2-D scatter plots of a UMAP layout, with the horizontal and vertical axes representing the two UMAP coordinates on continuous numeric scales. Panels (a2) and (b2) show corresponding 2-D projections from a tour of the 7-D data and model, where each axis represents a linear combination of the original seven dimensions. Data points are plotted as individual markers, and selected subsets are highlighted in purple while unselected points appear in a contrasting colour. In the top row, a set of points highlighted in the lower region of the UMAP layout in (a1) is linked to highlighted points in the tour view in (a2). In the bottom row, a different highlighted region in the UMAP layout in (b1) is linked to a different set of highlighted points in the tour view in (b2). The linked selections update consistently across panels, showing how the same subsets of observations appear in the UMAP layout and in different projections of the high-dimensional space."}

@@

```{r scurve-linkerror-html, echo=TRUE, eval=knitr::is_html_output(), fig.cap="Interactively linked plots connecting the residual distribution (left), the UMAP layout (middle) and the fitted model overlaid with the data in $7\\text{-}D$ (right) using `langevitour`. Brushing in any view highlights the corresponding subset across all panels, allowing detailed inspection of how residuals relate to both the UMAP embedding and the $p\\text{-}D$ structure. Points with higher residuals typically correspond to sparse or bridging regions in the UMAP layout and the less dense ends of the `scurve`, whereas low-residual points tend to cluster within denser, well-preserved regions. Double-clicking clears the selection and resets all highlights.", fig.alt = "The figure shows three interactively linked panels. The left panel displays a univariate plot of model residuals, with the horizontal axis representing residual values and the vertical axis representing density. The middle panel shows a 2-D UMAP layout of the scurve data, with horizontal and vertical axes representing the two UMAP coordinates. The right panel shows a 2-D projection of the 7-D data and fitted model using langevitour, where each axis represents a linear combination of the original dimensions. Points are plotted as individual markers in all panels, and brushing a subset of points in any panel highlights the corresponding points across the other two panels using a distinct colour. Selections update dynamically as the user interacts with any panel, and double-clicking clears all highlights. All panels maintain roughly square aspect ratios and consistent visual encodings for selected and unselected points throughout the interactive view.", layout = "l-body"}

@@

```{r scurve-linkerror-pdf, eval=knitr::is_latex_output(), fig.pos="H", fig.width=15, fig.height=10, fig.cap="Exploring residuals in relation to UMAP layouts using a $7\\text{-}D$ \\texttt{scurve} model. Three views are linked: distribution of residuals (a1, b1), UMAP layout (a2, b2), and projection of the $7\\text{-}D$ model with data (a3, b3). The purple points highlight selected subsets of the data, which differ across rows. In the top row (a1–a3), points with higher residuals (a1) are selected, corresponding to the sparse bridging region in the UMAP layout (a2) and the less dense end of the \\texttt{scurve} in the high-dimensional projection (a3). In the bottom row (b1–b3), points with lower residuals (b1) are highlighted, which map to one side of the dense region in the NLDR layout (b2) and to a thicker band of the \\texttt{scurve} in the projection (b3). This comparison illustrates how residuals can help diagnose distortions in UMAP, with high-residual points often concentrated in sparse or stretched regions of the structure.", fig.alt = "The figure consists of six panels arranged in two rows and three columns, showing residuals and their correspondence between UMAP and high-dimensional projections of the scurve data. In each row, the left panel (a1, b1) is a univariate plot of model residuals, with the horizontal axis representing residual values and the vertical axis representing frequency or density; points selected for highlighting appear in purple. The middle panel (a2, b2) shows a 2-D UMAP layout, with horizontal and vertical axes representing the two UMAP coordinates; the selected points from the left panel are highlighted in purple, while unselected points appear in a contrasting colour. The right panel (a3, b3) shows a 2-D projection of the 7-D data and model using a tour view, where axes represent linear combinations of the original dimensions; highlighted points correspond to those selected in the residuals and UMAP panels, while unselected points are shown in a contrasting colour. Across the panels, the purple highlights differ between the top and bottom rows, showing different subsets of points. All panels maintain roughly square aspect ratios and use consistent visual encodings for selection and non-selection."}

@@

In the implementation examples, points are shown without cluster-based colouring unless explicitly stated. When points are coloured by cluster, linked plot functionality is currently only partially supported: selections made in the `langevitour` controls are not reflected in the corresponding highlights in the interactive $2\text{-}D$ layout. As a result, cluster-specific exploration must be carried out separately in the tour view and the interactive $2\text{-}D$ layout. In addition to `langevitour`, linked plots can also be constructed using tour views generated with the `detourr` package. (see [vignette: 8. Linked plots with `detourr`](https://jayanilakshika.github.io/quollr/articles/quollr8linkeddetourr.html) for details.)

@@

```{r limb-hbe, fig.cap="Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using HBE for varying binwidth ($a_1$). Colour  used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d performs well at large binwidth (where the binwidth is not enough to capture the data structure) and poorly as the bin width decreases. Layout f is the best choice.\\label{fig:limb-hbe}", fig.alt = "A multi-panel figure shows five alternative 2D embeddings of the same high-dimensional data, arranged side by side (panels b–f). Each panel is a scatterplot with two unlabeled embedding axes (dimension 1 on the horizontal axis and dimension 2 on the vertical axis, each spanning a similar, roughly symmetric range around zero). Points represent the same set of observations in every panel, plotted with identical colour or marker scheme so that corresponding observations can be visually compared across layouts. Panel b shows a UMAP layout (neighbors = 5, minimum distance = 0.1), panel c a PHATE layout (knn = 5), panel d a TriMAP layout (inliers = 12, outliers = 4, random = 3), panel e a PaCMAP layout (neighbors = 10, init = random, MN-ratio = 0.5, FP-ratio = 2), and panel f a tSNE layout (perplexity = 15). Across panels, similar clusters of points appear but are arranged differently in space: some layouts show more clearly separated compact clusters, while others show overlapping or elongated groups, reflecting how each method balances local versus global structure. The accompanying HBE plot (typically a separate panel) summarizes model fit quality for each layout, with one value per method indicating that some embeddings provide a more faithful representation of the original data structure than others.", fig.pos="!ht", echo=TRUE}

@@

```{r model-limb, fig.cap="Compare the published $2\\text{-}D$ layout (Figure \\ref{fig:limb-hbe}a) and the $2\\text{-}D$ layout selected (Figure \\ref{fig:limb-hbe}f) by HBE plot (Figure \\ref{fig:limb-hbe}) from the tSNE, UMAP, PHATE, TriMAP, and PaCMAP with different hyper-parameters. The Limb muscle data ($n =  1067$) has seven close different shaped clusters in $10\\text{-}D$.", fig.alt = "The figure shows a 2-D scatter plot of a tSNE layout for the Limb muscle dataset (n=1067) with seven clusters in 10 dimensions. Each point represents one observation, with point colour indicating the cluster label reported in the original paper. The horizontal and vertical axes represent the two tSNE coordinates, both on roughly symmetric numeric scales around zero. Points form several visually distinct clusters, with some local variation visible within clusters. A model overlay with a binwidth of 0.06 shows the local density structure, providing a background grid that highlights differences between clusters without fragmenting the plot.", fig.pos="!ht", fig.width=15, fig.height=10, eval=knitr::is_latex_output()}

@@

```{r tsne-link-limb, eval=knitr::is_html_output(), fig.cap="Interactively linked plots connecting layout a (left) and the fitted model overlaid with the limb muscle data in $10\\text{-}D$ (right). Each color represents a different muscle group, with triangulated hexagon centroids on the NLDR layout connected to their corresponding regions in the high-dimensional projection using `langevitour`. Brushing in either view highlights the same subset of points across both panels, enabling exploration of how clusters in the low-dimensional embedding align with structures in the original high-dimensional space. Double-clicking clears the selection and removes highlights.", fig.alt = "The figure shows two interactively linked panels. The left panel displays a 2-D NLDR layout of the Limb muscle dataset (n=1067), with each point representing an observation and coloured according to its muscle group. The right panel shows a 2-D projection of the 10-D data and fitted model using langevitour, where each axis represents a linear combination of the original dimensions. Triangulated hexagon centroids on the NLDR layout are connected to their corresponding regions in the high-dimensional projection. Brushing a subset of points in either panel highlights the corresponding points in both panels using a distinct colour, and double-clicking clears all selections. Both panels maintain roughly square aspect ratios and consistent visual encodings for selected and unselected points throughout the interactive view.", layout = "l-body"}

@@

```{r tsne-link-limb-best, eval=knitr::is_html_output(), fig.cap="Interactively linked plots connecting layout f (left) and the fitted model overlaid with the limb muscle data in $10\\text{-}D$ (right). Each color represents a different muscle group, with triangulated hexagon centroids on the NLDR layout connected to their corresponding regions in the high-dimensional projection using `langevitour`. Brushing in either view highlights the same subset of points across both panels, enabling exploration of how clusters in the low-dimensional embedding align with structures in the original high-dimensional space. Double-clicking clears the selection and removes highlights.", fig.alt = "The figure shows two interactively linked panels. The left panel displays a 2-D NLDR layout (layout f) of the Limb muscle dataset (n=1067), with each point representing an observation and coloured by muscle group. The right panel shows a 2-D projection of the 10-D data and fitted model using langevitour, where each axis represents a linear combination of the original dimensions. Triangulated hexagon centroids on the NLDR layout are connected to their corresponding regions in the high-dimensional projection. Brushing a subset of points in either panel highlights the same points in both panels using a distinct colour, and double-clicking clears all selections. Both panels maintain roughly square aspect ratios and consistent visual encodings for selected and unselected points throughout the interactive view.", layout = "l-body"}

>>>>

@@

```{r geom-outputs, fig.cap="The outputs of `geom_hexgrid` and `geom_trimesh` include: (a) a complete hexagonal grid, (b) a hexagonal grid that corresponds with the data, (c) a full grid based on centroid triangulation, and (d) a centroid triangulation grid that aligns with the data.", fig.align='center', fig.pos="!ht", fig.height=5, fig.width=20, eval=knitr::is_html_output(), fig.alt = "The figure consists of four 2-D panels illustrating hexagonal grids and triangulated meshes. Panel (a) shows a complete hexagonal grid covering a rectangular 2-D region, with all hexagons drawn regardless of whether data are present. Panel (b) shows a hexagonal grid restricted to bins that contain data points, with only those hexagons displayed and gaps where bins are empty. Panel (c) shows a full triangulated mesh constructed from centroids across the entire grid region, with centroids connected by straight line segments to form triangles. Panel (d) shows a triangulated mesh constructed only from centroids corresponding to non-empty bins, aligned with the data layout and leaving open areas where no centroids are present. In all panels, the horizontal and vertical axes represent continuous 2-D coordinates with bounded numeric ranges, and the grids and meshes are drawn with uniform line styles and colors."}

@@

```{r geom-outputs-pdf, fig.cap="The outputs of \\texttt{geom\\_hexgrid} and \\texttt{geom\\_trimesh} include: (a) a complete hexagonal grid, (b) a hexagonal grid that corresponds with the data, (c) a full grid based on centroid triangulation, and (d) a centroid triangulation grid that aligns with the data.", fig.align='center', fig.pos="!ht", fig.height=5, fig.width=20, eval=knitr::is_latex_output(), fig.alt = "The figure consists of four 2-D panels illustrating hexagonal grids and triangulated meshes. Panel (a) shows a complete hexagonal grid covering a rectangular 2-D region, with all hexagons drawn regardless of whether data are present. Panel (b) shows a hexagonal grid restricted to bins that contain data points, with only those hexagons displayed and gaps where bins are empty. Panel (c) shows a full triangulated mesh constructed from centroids across the entire grid region, with centroids connected by straight line segments to form triangles. Panel (d) shows a triangulated mesh constructed only from centroids corresponding to non-empty bins, aligned with the data layout and leaving open areas where no centroids are present. In all panels, the horizontal and vertical axes represent continuous 2-D coordinates with bounded numeric ranges, and the grids and meshes are drawn with uniform line styles and colors."}

@@

```{r scurve-highd-model-pdf, eval=knitr::is_latex_output(), fig.pos="!ht", fig.cap="$2\\text{-}D$ projections of the lifted high-dimensional wireframe model from the \\texttt{scurve} UMAP layout. Each panel (a1–a4) shows the model (black) overlaid on \\texttt{scurve} data (blue) in different projections. These views illustrate how the lifted wireframe model captures the structure of the \\texttt{scurve} data. Regions with sparse or no data in the UMAP layout are also visible in the lifted model.", fig.alt = "The figure shows four panels of 2-D scatter plots (a1–a4), each displaying a different projection of the lifted high-dimensional wireframe model derived from the scurve UMAP layout. In each panel, the horizontal and vertical axes represent two projected coordinates with continuous numeric ranges. The wireframe model is drawn in black as a wireframe of connected line segments, and the scurve data points are overlaid in blue as individual markers. The relative positions and orientations of the wireframe and data points vary across panels as different projections are shown. In some regions, the wireframe extends into areas with few or no data points, leaving visible gaps between the model and the data. All panels have similar aspect ratios and use consistent color and symbol encodings to distinguish the model from the data."}

@@

```{r scurve-highd-model-detourrpdf, out.width=c("50%", "50%"), fig.cap="Screenshots of the lifted high-dimensional wireframe model from the \\texttt{scurve} UMAP layout using \\texttt{detourr}. Regions with sparse or no data in the UMAP layout are also visible in the lifted model.", fig.alt = "The figure shows several screenshots of 2-D projections of the lifted high-dimensional wireframe model derived from the scurve UMAP layout using detourr. In each screenshot, the horizontal and vertical axes represent projected coordinates with continuous numeric ranges. The wireframe model is displayed as a set of connected line segments, and the scurve data points are shown as individual markers overlaid on the wireframe. The orientation and shape of the wireframe and point cloud differ across screenshots as different projections are shown. In some views, parts of the wireframe extend into areas with few or no data points, leaving visible gaps between the model and the data. All screenshots use consistent color and line styles and have similar aspect ratios.", fig.pos="H", fig.show='hold', eval=knitr::is_latex_output()}

@@

```{r scurve-nldrlink-html, echo=TRUE, eval=knitr::is_html_output(), fig.cap="Interactively linked plots connecting the UMAP layout (left) and the fitted model overlaid with the data in $7\\text{-}D$ (right) using `langevitour`. Brushing in either view highlights the corresponding subset of points across both representations, allowing examination of how well the UMAP layout preserves the underlying `scurve` structure. Double-clicking on either view clears the selection and dehighlights all points.", fig.alt = "The figure shows two interactively linked panels. The left panel displays a 2-D UMAP layout of the scurve data, with horizontal and vertical axes representing the UMAP coordinates on continuous numeric scales. The right panel shows a 2-D projection of the 7-D data and fitted model using langevitour, where each axis represents a linear combination of the original seven dimensions. Data points are plotted as individual markers, and brushing a subset of points in either panel highlights the corresponding points in the other panel using a distinct color. The highlighted points update dynamically as selections change, and double-clicking on either panel clears all highlights. The display maintains roughly square aspect ratios in both panels and uses consistent visual encodings for data and selected subsets throughout the interactive view.", layout = "l-body"}

@@

```{r scurve-nldrlink-pdf, eval=knitr::is_latex_output(), fig.pos="!ht", fig.width=10, fig.height=10, fig.cap="Exploring the correspondence between UMAP layout and \\texttt{scurve} structure in $7\\text{-}D$. Two sets of plots are interactively linked: UMAP layout (a1, b1) and projection of $7\\text{-}D$ model and data (a2, b2). The purple points indicate the selected subsets, which differ between rows. In (a1), the lower bridge of the \\texttt{scurve} is highlighted, which corresponds in (a2) to points spanning across both arms of the high-dimensional structure. In (b1), a different region near the upper arm of the \\texttt{scurve} is selected, and in (b2) these points map onto one side of the curved manifold in $7\\text{-}D$ projection. While the UMAP layout suggests distinct local clusters, the linked tour views reveal how these selections trace continuous structures in the $7\\text{-}D$ space, highlighting distortions introduced by UMAP.", fig.alt = "The figure consists of four linked panels arranged in two rows and two columns. Panels (a1) and (b1) show 2-D scatter plots of a UMAP layout, with the horizontal and vertical axes representing the two UMAP coordinates on continuous numeric scales. Panels (a2) and (b2) show corresponding 2-D projections from a tour of the 7-D data and model, where each axis represents a linear combination of the original seven dimensions. Data points are plotted as individual markers, and selected subsets are highlighted in purple while unselected points appear in a contrasting color. In the top row, a set of points highlighted in the lower region of the UMAP layout in (a1) is linked to highlighted points in the tour view in (a2). In the bottom row, a different highlighted region in the UMAP layout in (b1) is linked to a different set of highlighted points in the tour view in (b2). The linked selections update consistently across panels, showing how the same subsets of observations appear in the UMAP layout and in different projections of the high-dimensional space."}

@@

```{r scurve-linkerror-html, echo=TRUE, eval=knitr::is_html_output(), fig.cap="Interactively linked plots connecting the residual distribution (left), the UMAP layout (middle) and the fitted model overlaid with the data in $7\\text{-}D$ (right) using `langevitour`. Brushing in any view highlights the corresponding subset across all panels, allowing detailed inspection of how residuals relate to both the UMAP embedding and the $p\\text{-}D$ structure. Points with higher residuals typically correspond to sparse or bridging regions in the UMAP layout and the less dense ends of the `scurve`, whereas low-residual points tend to cluster within denser, well-preserved regions. Double-clicking clears the selection and resets all highlights.", fig.alt = "The figure shows three interactively linked panels. The left panel displays a univariate plot of model residuals, with the horizontal axis representing residual values and the vertical axis representing density. The middle panel shows a 2-D UMAP layout of the scurve data, with horizontal and vertical axes representing the two UMAP coordinates. The right panel shows a 2-D projection of the 7-D data and fitted model using langevitour, where each axis represents a linear combination of the original dimensions. Points are plotted as individual markers in all panels, and brushing a subset of points in any panel highlights the corresponding points across the other two panels using a distinct color. Selections update dynamically as the user interacts with any panel, and double-clicking clears all highlights. All panels maintain roughly square aspect ratios and consistent visual encodings for selected and unselected points throughout the interactive view.", layout = "l-body"}

@@

```{r scurve-linkerror-pdf, eval=knitr::is_latex_output(), fig.pos="H", fig.width=15, fig.height=10, fig.cap="Exploring residuals in relation to UMAP layouts using a $7\\text{-}D$ \\texttt{scurve} model. Three views are linked: distribution of residuals (a1, b1), UMAP layout (a2, b2), and projection of the $7\\text{-}D$ model with data (a3, b3). The purple points highlight selected subsets of the data, which differ across rows. In the top row (a1–a3), points with higher residuals (a1) are selected, corresponding to the sparse bridging region in the UMAP layout (a2) and the less dense end of the \\texttt{scurve} in the high-dimensional projection (a3). In the bottom row (b1–b3), points with lower residuals (b1) are highlighted, which map to one side of the dense region in the NLDR layout (b2) and to a thicker band of the \\texttt{scurve} in the projection (b3). This comparison illustrates how residuals can help diagnose distortions in UMAP, with high-residual points often concentrated in sparse or stretched regions of the structure.", fig.alt = "The figure consists of six panels arranged in two rows and three columns, showing residuals and their correspondence between UMAP and high-dimensional projections of the scurve data. In each row, the left panel (a1, b1) is a univariate plot of model residuals, with the horizontal axis representing residual values and the vertical axis representing frequency or density; points selected for highlighting appear in purple. The middle panel (a2, b2) shows a 2-D UMAP layout, with horizontal and vertical axes representing the two UMAP coordinates; the selected points from the left panel are highlighted in purple, while unselected points appear in a contrasting color. The right panel (a3, b3) shows a 2-D projection of the 7-D data and model using a tour view, where axes represent linear combinations of the original dimensions; highlighted points correspond to those selected in the residuals and UMAP panels, while unselected points are shown in a contrasting color. Across the panels, the purple highlights differ between the top and bottom rows, showing different subsets of points. All panels maintain roughly square aspect ratios and use consistent visual encodings for selection and non-selection."}

@@

In the implementation examples, points are shown without cluster-based coloring unless explicitly stated. When points are colored by cluster, linked plot functionality is currently only partially supported: selections made in the `langevitour` controls are not reflected in the corresponding highlights in the interactive $2\text{-}D$ layout. As a result, cluster-specific exploration must be carried out separately in the tour view and the interactive $2\text{-}D$ layout. In addition to `langevitour`, linked plots can also be constructed using tour views generated with the `detourr` package. (see [vignette: 8. Linked plots with `detourr`](https://jayanilakshika.github.io/quollr/articles/quollr8linkeddetourr.html) for details.)

@@

```{r limb-hbe, fig.cap="Assessing which of the 6 NLDR layouts on the limb muscle data is the better representation using HBE for varying binwidth ($a_1$). Color used for the lines and points in the left plot and in the scatterplots represents NLDR layout (a-f). Layout d performs well at large binwidth (where the binwidth is not enough to capture the data structure) and poorly as the bin width decreases. Layout f is the best choice.\\label{fig:limb-hbe}", fig.alt = "A multi-panel figure shows five alternative 2D embeddings of the same high-dimensional data, arranged side by side (panels b–f). Each panel is a scatterplot with two unlabeled embedding axes (dimension 1 on the horizontal axis and dimension 2 on the vertical axis, each spanning a similar, roughly symmetric range around zero). Points represent the same set of observations in every panel, plotted with identical color or marker scheme so that corresponding observations can be visually compared across layouts. Panel b shows a UMAP layout (neighbors = 5, minimum distance = 0.1), panel c a PHATE layout (knn = 5), panel d a TriMAP layout (inliers = 12, outliers = 4, random = 3), panel e a PaCMAP layout (neighbors = 10, init = random, MN-ratio = 0.5, FP-ratio = 2), and panel f a tSNE layout (perplexity = 15). Across panels, similar clusters of points appear but are arranged differently in space: some layouts show more clearly separated compact clusters, while others show overlapping or elongated groups, reflecting how each method balances local versus global structure. The accompanying HBE plot (typically a separate panel) summarizes model fit quality for each layout, with one value per method indicating that some embeddings provide a more faithful representation of the original data structure than others.", fig.pos="!ht", echo=TRUE}

@@

```{r model-limb, fig.cap="Representative views of two selected NLDR layouts for the Limb muscle dataset ($n=1067$), shown row-wise. The top row (a1–a3) corresponds to the published $2\\text{-}D$ layout (Figure \\ref{fig:limb-hbe}a), and the bottom row (f1–f3) corresponds to the $2\\text{-}D$ layout selected (Figure \\ref{fig:limb-hbe}f) selected using the HBE plot. In each row, the left panel (a1, f1) shows the NLDR embedding with points colored by muscle group and overlaid with triangulated hexagon centroids. The middle (a2, f2) and right (a3, f3) panels show two different $2\\text{-}D$ projections of the fitted model and data in the original $10\\text{-}D$ space, with the same triangular mesh displayed. Together, these panels summarize how the low-dimensional layouts relate to the underlying high-dimensional structure across different viewing directions.", fig.alt = "The figure consists of six panels arranged in two rows and three columns, showing representative views of two NLDR layouts for the Limb muscle dataset (n=1067). In the top row (a1–a3), the left panel (a1) shows the published 2-D NLDR embedding with points colored by muscle group and overlaid with triangulated hexagon centroids. The middle (a2) and right (a3) panels show two different 2-D projections of the corresponding 10-D data and fitted model, with points plotted as individual markers and the same triangular mesh connecting centroids. In the bottom row (f1–f3), the left panel (f1) shows the 2-D NLDR layout selected using the HBE plot, again with colored points and triangulated centroids, while panels (f2) and (f3) show two different 2-D projections of the associated 10-D structure with the same mesh displayed. Across all panels, axes represent continuous numeric coordinates, color encodes muscle group, and the triangular mesh is used consistently to show correspondence between the low-dimensional layouts and the high-dimensional projections.", fig.pos="!ht", fig.width=15, fig.height=10, eval=knitr::is_latex_output()}

@@

```{r tsne-link-limb, eval=knitr::is_html_output(), fig.cap="Interactively linked plots connecting layout a (left) and the fitted model overlaid with the limb muscle data in $10\\text{-}D$ (right). Each color represents a different muscle group, with triangulated hexagon centroids on the NLDR layout connected to their corresponding regions in the high-dimensional projection using `langevitour`. Brushing in either view highlights the same subset of points across both panels, enabling exploration of how clusters in the low-dimensional embedding align with structures in the original high-dimensional space. Double-clicking clears the selection and removes highlights.", fig.alt = "The figure shows two interactively linked panels. The left panel displays a 2-D NLDR layout of the Limb muscle dataset (n=1067), with each point representing an observation and colored according to its muscle group. The right panel shows a 2-D projection of the 10-D data and fitted model using langevitour, where each axis represents a linear combination of the original dimensions. Triangulated hexagon centroids on the NLDR layout are connected to their corresponding regions in the high-dimensional projection. Brushing a subset of points in either panel highlights the corresponding points in both panels using a distinct color, and double-clicking clears all selections. Both panels maintain roughly square aspect ratios and consistent visual encodings for selected and unselected points throughout the interactive view.", layout = "l-body"}

@@

```{r tsne-link-limb-best, eval=knitr::is_html_output(), fig.cap="Interactively linked plots connecting layout f (left) and the fitted model overlaid with the limb muscle data in $10\\text{-}D$ (right). Each color represents a different muscle group, with triangulated hexagon centroids on the NLDR layout connected to their corresponding regions in the high-dimensional projection using `langevitour`. Brushing in either view highlights the same subset of points across both panels, enabling exploration of how clusters in the low-dimensional embedding align with structures in the original high-dimensional space. Double-clicking clears the selection and removes highlights.", fig.alt = "The figure shows two interactively linked panels. The left panel displays a 2-D NLDR layout (layout f) of the Limb muscle dataset (n=1067), with each point representing an observation and colored by muscle group. The right panel shows a 2-D projection of the 10-D data and fitted model using langevitour, where each axis represents a linear combination of the original dimensions. Triangulated hexagon centroids on the NLDR layout are connected to their corresponding regions in the high-dimensional projection. Brushing a subset of points in either panel highlights the same points in both panels using a distinct color, and double-clicking clears all selections. Both panels maintain roughly square aspect ratios and consistent visual encodings for selected and unselected points throughout the interactive view.", layout = "l-body"}
